\section*{Disclaimer}
\addcontentsline{toc}{subsubsection}{Disclaimer}

We ask that you (our peer) respect all of the guidelines we (your peers) set forth with respect to these review notes. If you are unable to respect any of the guidelines mentioned in the proceeding few paragraphs, please delete this file or close the hard copy of the review notes.\\

\begin{enumerate}

\item We alone reserve the right to alter and distribute these review notes. If you see a mistake or feel as if important content is missing, contact us (information provided below) so that we can make the appropriate correction(s).

\item We will be supporting the accuracy of these review notes for the remainder of the summer semester (2013). After such a time these review notes should be considered obsolete.

\item These review notes are provided absolutely free of charge. If you paid for a hard copy or e-copy of these review notes then you are obligated to contact us so that we can take the appropriate action(s) towards the distributor.

\end{enumerate}

Keep in mind that these notes are in no way guaranteed to be accurate. There may be mistakes, outdated information or tangents that will not be directly related to some of the material presented in the course. These notes have been developed by a few of your peers (yes, students!) during our undergraduate year taking CS 136. No professor, TA or anyone involved in the administration at the University of Waterloo has endorsed these notes.

\section*{Special Notices}
\addcontentsline{toc}{subsubsection}{Special Notices}

These review notes would not be possible without the offering of CS 136 at the University of Waterloo and all the hard work the professors, TAs and behind the scenes administration put into this course. This is a truly fantastic course and every student involved in creating these review notes are absolutely loving it. Special mentions to the following students:

\begin{itemize}

\item Jacob Pollack, author (\textbf{jpollack@uwaterloo.ca}).
\item Jacob Willemsma, co-author (\textbf{wjwillem@uwaterloo.ca}).

\end{itemize}

\subsubsection*{Acknowledgements}

Some further acknowledgments to the authors of the CS 136 material and other influencing figures.

\begin{itemize}

\item Professors and Teaching Assistants for the summer semester of 2013: Dave Tompkins, Olga Zorin and Patrick Nicholson.
\item Design inspired by "Linear Algebra Course Notes" by Dan Wolczuk (with permission).

\end{itemize}
